Benchmarking FDR Methods - Summarizing Simulation Results (Part II)

### 2017-11-06 datasets/simulations/analysis-04-plot-psim-summaries.Rmd
Add initial analysis of psim results

---
title: "Benchmarking FDR Methods - Summarizing Simulation Results"
author: "Rafalab Journal Club Members"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    number_sections: true
    code_folding: hide
---

@@

```{r, process-data, message=FALSE}
## list of results from 100 replication experiments
dl <- list.dirs("data-tsim/M100", recursive=FALSE, full.names=TRUE)

@@

In this analysis, we look at summarizing the results for the various simulation settings considered under the two-sample t-test setup. Each simulation included 20,000 two-sample t-tests performed over a variable number of null (group difference 0) and alternative (group difference > 0) datasets. Different simulation settings were considered to evaluate the impact of varying sample sizes, effect sizes, effect size distributions, proportions of null hypotheses, and presence or absence of an informative covariate. More details on the simulation settings considered are included in the header of the `tsim-core-settings.R` script. Each setting was replicated 100 times.

In a previous analysis (`analysis-01-plot.Rmd`), each simulation setting was considered and plotted separately with performance metrics reported over a range of FDR ("alpha") thresholds. However, since the total number of settings considered is quite large, processing and interpreting these results was both cumbersome and difficult. To get a higher-level picture of the simulation results, here we consider a single FDR threshold, `alpha = 0.05`, and plot performance metrics across settings for each method. Plots for FPR and "rejectprop" (rejection proportion) include blue horizontal lines at the expected FDR level (`0.05`) and proportion of alternative hypotheses (`1 - pi0`), respectively.

For simulation settings where the effect size (expected group differece) is not varying, a fixed difference of `1.5` was used for all alternative hypotheses. For simulation settings with the uniform (un-informative) or SE informative covariate, if the proportion of null hypotheses is not varying, `80%` of hypotheses were simulated under the null. For simulation settings with the "Boca-Leek" style informative covariate, a functional relationship was first defined between the probability of a test being null and the "informative covariate". Then, the informative covariates were sampled from Uniform(0, 1) and used to randomly assign each test as null or alternative according to the predefined functional relationship. The form of the relationship used here was taken from Setting II of the Boca-Leek simulations.

@@

```{r}
lab_order <- c("Bonferroni", "Benjamini-Hochberg", "q-value", "local FDR",
               "IHW", "Boca-Leek", "FDRreg (theoretical)", "FDRreg (empirical)",
               "ASH", "unadjusted")
```

@@

```{r}
re_blabel <- function(tab) {
    tsb %>%
        mutate(blabel = fct_recode(f = blabel, "IHW" = "ihw-a05", "ASH" = "ashs",
                                   "Boca-Leek" = "bl-df03", "local FDR" = "lfdr",
                                   "FDRreg (theoretical)" = "scott-theoretical",
                                   "FDRreg (empirical)" = "scott-empirical",
                                   "Benjamini-Hochberg" = "bh",
                                   "Bonferroni" = "bonf", "q-value" = "qvalue"),
               takes_cov = blabel %in%
                   c("IHW", "ASH", "Boca-Leek", "FDRreg (theoretical)", "FDRreg (empirical)")) %>%
        mutate(blabel = factor(blabel, levels = lab_order))
}
```

@@

```{r}
f <- "data-tsim/M100/esize_fixed-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Effect Size, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
### SE as Covariate

```{r}
f <- "data-tsim/M100/esize_fixed-se/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Effect Size, SE as Covariate (100 replications)",
         tsb, cutoffs)
```

@@

```{r}
f <- "data-tsim/M100/esize_fixed-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Effect Size, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```

@@

@@

```{r}
f <- "data-tsim/M100/pi0-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Null Proportion, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
### SE as Covariate

```{r}
f <- "data-tsim/M100/pi0-se/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Null Proportion, SE as Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
## Varying Sample Sizes
<!-- ######################################################################## -->

@@

```{r}
f <- "data-tsim/M100/n-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## can also add default of n = 10 from alt pi0 simulation set
## (same effect_size = 1.5, pi0 = 0.8)
f2 <- "data-tsim/M100/pi0-uniform/results-performanceMetrics-alpha05.rds"
tsb2 <- readRDS(f2)
tsb2 %<>%
    filter(param == "results-altp0_8-M100") %>%
    mutate(param = "results-altn_10-M100")

tsb <- bind_rows(tsb, tsb2)

## remove rejections: scaling is off from rest (also dup'd w/ rejprop)
## remove FWER: not informative as a boxplot
tsb %<>%
    mutate(param = gsub(".*?-.*?_(.*?)-.*", "\\1", param) %>% as.numeric,
           performanceMetric =
               factor(performanceMetric,
                      levels = c("rejections", "rejectprop", "TPR",
                                 "TNR", "FPR", "FNR", "FWER"))) %>%
    filter(!(performanceMetric %in% c("rejections", "FWER"))) %>%
    mutate(param = factor(paste0("n (per group) = ", param),
                          levels = paste0("n (per group) = ", c(3, 6, 10, 20))))

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Sample Sizes, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
### SE as Covariate

```{r}
f <- "data-tsim/M100/n-se/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## can also add default of n = 10 from alt pi0 simulation set
## (same effect_size = 1.5, pi0 = 0.8)
f2 <- "data-tsim/M100/pi0-se/results-performanceMetrics-alpha05.rds"
tsb2 <- readRDS(f2)
tsb2 %<>%
    filter(param == "results-altp0_8-M100") %>%
    mutate(param = "results-altn_10-M100")

tsb <- bind_rows(tsb, tsb2)

## remove rejections: scaling is off from rest (also dup'd w/ rejprop)
## remove FWER: not informative as a boxplot
tsb %<>%
    mutate(param = gsub(".*?-.*?_(.*?)-.*", "\\1", param) %>% as.numeric,
           performanceMetric =
               factor(performanceMetric,
                      levels = c("rejections", "rejectprop", "TPR",
                                 "TNR", "FPR", "FNR", "FWER"))) %>%
    filter(!(performanceMetric %in% c("rejections", "FWER"))) %>%
    mutate(param = factor(paste0("n (per group) = ", param),
                          levels = paste0("n (per group) = ", c(3, 6, 10, 20))))

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Sample Sizes, SE as Covariate (100 replications)",
         tsb, cutoffs)
```

```{r}
f <- "data-tsim/M100/n-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## remove rejections: scaling is off from rest (also dup'd w/ rejprop)
## remove FWER: not informative as a boxplot
tsb %<>%
    mutate(param = gsub(".*?-.*?_(.*?)-.*", "\\1", param) %>% as.numeric,
           performanceMetric =
               factor(performanceMetric,
                      levels = c("rejections", "rejectprop", "TPR",
                                 "TNR", "FPR", "FNR", "FWER"))) %>%
    filter(!(performanceMetric %in% c("rejections", "FWER"))) %>%
    mutate(param = factor(paste0("n (per group) = ", param),
                          levels = paste0("n (per group) = ", c(3, 6, 20))))

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Sample Sizes, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
## Randomly-Sampled Effect Size
<!-- ######################################################################## -->

```{r}
f <- "data-tsim/M100/esize_random-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Randomly Sampled Effect, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
### SE as Covariate

```{r}
f <- "data-tsim/M100/esize_random-se/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Randomly Sampled Effect, SE as Covariate (100 replications)",
         tsb, cutoffs)
```

@@

```{r}
f <- "data-tsim/M100/esize_random-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Two-Sample T-Tests w/ Varying Randomly Sampled Effect, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```

>>>>

---
title: "Benchmarking FDR Methods - Summarizing Simulation Results (Part II)"
author: "Rafalab Journal Club Members"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    number_sections: true
    code_folding: hide
---

@@

```{r, process-data, message=FALSE}
## list of results from 100 replication experiments
dl <- list.dirs("data-psim/M100", recursive=FALSE, full.names=TRUE)

@@

This analysis is similar to the one presented in `analysis-03-plot-tsim-summaries.[Rmd|html]`. However, rather than looking at the results of two-sample t-tests, here, we consider simulations performed directly at the level of effect sizes and p-values without any underlying data. This approach to simulating data follows similar simulations carried out in the Boca-Leek, IHW ("size investing strategy" simulations), and ASH papers. Different simulation settings were considered to evaluate the impact of varying effect sizes, effect size distributions, effect size sampling distributions, proportions of null hypotheses, and presence or absence of an informative covariate. More details on the simulation settings considered are included in the header of the `psim-core-settings.R` script. Each setting was replicated 100 times.

Since these simulations do not rely on underlying simulated data, but rather simulate summary statistics directly, there is no way of calculating a corresponding standard error (SE) for each hypothesis test. *As a result, the ASH method, which requires the SE for each test, is not included in these simulations.*

As in `analysis-03-plot-tsim-summaries.[Rmd|html]`, simulation results are reported across settings for each method at the `alpha = 0.05` FDR threshold.

@@

```{r}
lab_order <- c("Bonferroni", "Benjamini-Hochberg", "q-value", "local FDR",
               "IHW", "Boca-Leek", "FDRreg (theoretical)", "FDRreg (empirical)",
               "unadjusted")
```

@@

```{r}
re_blabel <- function(tab) {
    tsb %>%
        mutate(blabel = fct_recode(f = blabel, "IHW" = "ihw-a05",
                                   "Boca-Leek" = "bl-df03", "local FDR" = "lfdr",
                                   "FDRreg (theoretical)" = "scott-theoretical",
                                   "FDRreg (empirical)" = "scott-empirical",
                                   "Benjamini-Hochberg" = "bh",
                                   "Bonferroni" = "bonf", "q-value" = "qvalue"),
               takes_cov = blabel %in%
                   c("IHW", "Boca-Leek", "FDRreg (theoretical)", "FDRreg (empirical)")) %>%
        mutate(blabel = factor(blabel, levels = lab_order))
}
```

@@

```{r}
f <- "data-psim/M100/esize_fixed-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Effect Size, Non-Informative Covariate (100 replications)",
         tsb, cutoffs) 
```

@@

```{r}
f <- "data-psim/M100/esize_fixed-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Effect Size, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```

@@

Need to re-run these simulations with larger test statistic value under the alternative - TPR is too low to be particularly interesting/informative for comparing across methods.

@@

```{r}
## older run with alternative tests at 1.5.
f <- "data-psim/M100/_pi0-uniform-delta15/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Null Proportion, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
## Randomly-Sampled UA Effect Size
<!-- ######################################################################## -->

Need to re-run these simulations with lower sampling variance - TPR is too low to be particularly interesting/informative for comparing across methods.

@@

```{r}
f <- "data-psim/M100/esize_random_ua-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Randomly Sampled UA Effect, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

```{r}
f <- "data-psim/M100/esize_random_ua-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Randomly Sampled UA Effect, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
## Randomly-Sampled Mean-Shift Effect Size
<!-- ######################################################################## -->

Chi-sq simulations look reasonable, but probably need to re-run the normal and t-test simulations with larger test statistic value under the alternative (possibly 3) - TPR is too low with shift of 1 and 2 to be particularly interesting/informative for comparing across methods.

```{r}
f <- "data-psim/M100/esize_random_shift-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Randomly Sampled Mean Shifted Effect, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

@@

```{r}
f <- "data-psim/M100/esize_random_shift-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

@@

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Randomly Sampled Mean Shifted Effect, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```

<!-- ######################################################################## -->
## Varying Parameter Sampling Distribution
<!-- ######################################################################## -->

Similar as the simulation settings considered above: chi-sq simulations look reasonable, but probably need to re-run the normal and t-test simulations with larger test statistic value under the alternative (possibly 3) - TPR is too low with shift of 1 and 2 to be particularly interesting/informative for comparing across methods.

```{r}
f <- "data-psim/M100/altnoise-uniform/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Sampling Distribution, Non-Informative Covariate (100 replications)",
         tsb, cutoffs)
```

```{r}
f <- "data-psim/M100/altnoise-bl/results-performanceMetrics-alpha05.rds"
tsb <- readRDS(f)

## summarize results in plot
gen_plot("Direct Parameter Simulation w/ Varying Sampling Distribution, Boca-Leek Style Covariate (100 replications)",
         tsb, cutoffs)
```
